---
title: "Hayman Fire Recovery"
author: "Keian Freshwater"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

**Library set-up for the document:**
```{r setup, warning=F,message=F}
library(tidyverse)
library(tidyr)
library(ggthemes)
library(lubridate)

# Now that we have learned how to munge (manipulate) data
# and plot it, we will work on using these skills in new ways

knitr::opts_knit$set(root.dir='..')
```

**Reading in Selected Datasets:**
```{r dataread, warning=F,message=F}
####-----Reading in Data and Stacking it ----- ####
#Reading in files
files <- list.files('data',full.names=T)


#Read in individual data files
ndmi <- read_csv(files[1]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')


ndsi <- read_csv(files[2]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')

ndvi <- read_csv(files[3])%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')

# Stack as a tidy dataset
full_long <- rbind(ndvi,ndmi,ndsi) %>%
  gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value))


```




## Question 1 

### Correlation between NDVI and NDMI, using a wide dataset. Data is between 1984-2019, and focuses on summer months (Jun-Sept):

```{r, warning=F}
full_wide <- spread(full_long,data,value) %>%
    mutate(month=month(DateTime),
           year=year(DateTime)) %>%
    filter(month %in% c(6,7,8,9))
```
**Plotting the Correlation between NDVI and NDMI:**
```{r,warning=F}
ggplot(data=full_wide,aes(x=ndmi,y=ndvi, color=month)) +
       geom_point() +
       xlab('NDMI') +
       ylab ('NDVI')
         
```

  
## Question 2 

### Correlation between average NDSI (normalized snow index) for January - April and average NDVI for June-August. Does the previous year's snow cover influence vegetation growth for the following summer?

**Creating a clean data set containing relevant averages:**
```{r, warning=F, message=F}
full_wide_full <- spread(full_long,data,value) %>%
  mutate(month=month(DateTime),
           year=year(DateTime)) %>%
  group_by(site,year)

```
```{r, warning=F, message=F}
full_wide_average_summer <- full_wide_full  %>%
  filter(month %in% c(6,7,8))%>%
  group_by(year) %>%
  summarize(mean_NDVI_summer = mean(ndvi,na.rm=TRUE))
```
```{r, warning=F, message=F}
full_wide_average_spring <- full_wide_full  %>%
  filter(month %in% c(1,2,3,4))%>%
  group_by(year) %>%
  summarize(mean_NDSI_spring = mean(ndsi,na.rm=TRUE))
```
```{r, warning=F, message=F}
full_wide_average <- inner_join(full_wide_average_spring,full_wide_average_summer)
```
**Plotting the Correlation between average summer NDVI and average spring NDSI:**
```{r, warning=F, message=F}
ggplot(data=full_wide_average,aes(x=mean_NDSI_spring,y=mean_NDVI_summer, color=year)) +
       geom_point() +
       xlab('Average NDSI (Jan-Apr)') +
       ylab ('Average NDVI (Jun-Aug)')
```

  
## Q3

### How is the snow effect from question 2 different between pre- and post-burn and burned and unburned? 

**Filter by Burned/Unburned:**
```{r, warning=F, message=F}
full_wide_unburned <- full_wide_full  %>%
  filter(site %in% c('unburned'))
full_wide_burned <- full_wide_full  %>%
  filter(site %in% c('burned'))
```
**Filter and Graph Unburned average summer NDVI and average spring NDSI:**
```{r, warning=F, message=F}
full_wide_unburned_average_summer <- full_wide_unburned  %>%
  filter(month %in% c(6,7,8))%>%
  group_by(year) %>%
  summarize(mean_NDVI_summer = mean(ndvi,na.rm=TRUE))
```
```{r, warning=F, message=F}
full_wide_unburned_average_spring <- full_wide_unburned  %>%
  filter(month %in% c(1,2,3,4))%>%
  group_by(year) %>%
  summarize(mean_NDSI_spring = mean(ndsi,na.rm=TRUE))
```
```{r, warning=F, message=F}
full_wide_average_unburned <- inner_join(full_wide_unburned_average_spring,full_wide_unburned_average_summer)
```
```{r, warning=F, message=F}
ggplot(data=full_wide_average_unburned,aes(x=mean_NDSI_spring,y=mean_NDVI_summer, color=year)) +
       geom_point() +
       xlab('Average NDSI (Jan-Apr)') +
       ylab ('Average NDVI (Jun-Aug)')
```
  
**Filter and Graph Burned average summer NDVI and average spring NDSI:**
```{r, warning=F, message=F}
full_wide_burned_average_summer <- full_wide_burned  %>%
  filter(month %in% c(6,7,8))%>%
  group_by(year) %>%
  summarize(mean_NDVI_summer = mean(ndvi,na.rm=TRUE))
```
```{r, warning=F, message=F}
full_wide_burned_average_spring <- full_wide_burned  %>%
  filter(month %in% c(1,2,3,4))%>%
  group_by(year) %>%
  summarize(mean_NDSI_spring = mean(ndsi,na.rm=TRUE))
```
```{r, warning=F, message=F}
full_wide_average_burned <- inner_join(full_wide_burned_average_summer,full_wide_burned_average_spring)
```
```{r, warning=F, message=F}
ggplot(data=full_wide_average_burned,aes(x=mean_NDSI_spring,y=mean_NDVI_summer, color=year)) +
       geom_point() +
       xlab('Average NDSI (Jan-Apr)') +
       ylab ('Average NDVI (Jun-Aug)')
```

  
## Question 4

**For the following questions:**
```{r, warning=F, message=F}
full_wide_jan <- full_wide_full  %>%
  filter(month %in% c(1)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_feb <- full_wide_full  %>%
  filter(month %in% c(2)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_mar <- full_wide_full  %>%
  filter(month %in% c(3)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_apr <- full_wide_full  %>%
  filter(month %in% c(4)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_may <- full_wide_full  %>%
  filter(month %in% c(5)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_jun <- full_wide_full  %>%
  filter(month %in% c(6)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_jul <- full_wide_full  %>%
  filter(month %in% c(7)) %>%
  select(-DateTime) %>%
  group_by(year,ndvi,ndmi,ndsi)
full_wide_aug <- full_wide_full  %>%
  filter(month %in% c(8)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_sept <- full_wide_full  %>%
  filter(month %in% c(9)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_oct <- full_wide_full  %>%
 filter(month %in% c(10)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_nov <- full_wide_full  %>%
  filter(month %in% c(11)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
full_wide_dec <- full_wide_full  %>%
  filter(month %in% c(12)) %>%
  group_by(month) %>%
  select(-DateTime,-site,-year)
```


### What month is the greenest month on average? 

**Calculate Averages for NDVI:**

```{r}
jan_NDVI <- mean(full_wide_jan$ndvi,na.rm=TRUE)
feb_NDVI <- mean(full_wide_feb$ndvi,na.rm=TRUE)
mar_NDVI <- mean(full_wide_mar$ndvi,na.rm=TRUE)
apr_NDVI <- mean(full_wide_apr$ndvi,na.rm=TRUE)
may_NDVI <- mean(full_wide_may$ndvi,na.rm=TRUE)
jun_NDVI <- mean(full_wide_jun$ndvi,na.rm=TRUE)
jul_NDVI <- mean(full_wide_jul$ndvi,na.rm=TRUE)
aug_NDVI <- mean(full_wide_aug$ndvi,na.rm=TRUE)
sept_NDVI <- mean(full_wide_sept$ndvi,na.rm=TRUE)
oct_NDVI <- mean(full_wide_oct$ndvi,na.rm=TRUE)
nov_NDVI <- mean(full_wide_nov$ndvi,na.rm=TRUE)
dec_NDVI <- mean(full_wide_nov$ndvi,na.rm=TRUE)
```
**Organize in Table:**
```{r}
green_NVDI <- rbind.data.frame(jan_NDVI,feb_NDVI,mar_NDVI,apr_NDVI,may_NDVI,jun_NDVI,jul_NDVI,aug_NDVI,sept_NDVI,oct_NDVI,nov_NDVI,dec_NDVI) %>%
  rename(NDVI = c.0.218601234567901..0.198653947368421..0.2019..0.250171647509579..)
greenest_month <- max(green_NVDI)
```
The greenest month is August, with an average NDVI of `r greenest_month`.


## Question 5
### What month is the snowiest on average?

**Calculate Averages for NDVI:**

```{r}
jan_NDSI <- mean(full_wide_jan$ndsi,na.rm=TRUE)
feb_NDSI <- mean(full_wide_feb$ndsi,na.rm=TRUE)
mar_NDSI <- mean(full_wide_mar$ndsi,na.rm=TRUE)
apr_NDSI <- mean(full_wide_apr$ndsi,na.rm=TRUE)
may_NDSI <- mean(full_wide_may$ndsi,na.rm=TRUE)
jun_NDSI <- mean(full_wide_jun$ndsi,na.rm=TRUE)
jul_NDSI <- mean(full_wide_jul$ndsi,na.rm=TRUE)
aug_NDSI <- mean(full_wide_aug$ndsi,na.rm=TRUE)
sept_NDSI <- mean(full_wide_sept$ndsi,na.rm=TRUE)
oct_NDSI <- mean(full_wide_oct$ndsi,na.rm=TRUE)
nov_NDSI <- mean(full_wide_nov$ndsi,na.rm=TRUE)
dec_NDSI <- mean(full_wide_nov$ndsi,na.rm=TRUE)
```
**Organize in Table:**
```{r}
snow_NDSI <- rbind.data.frame(jan_NDSI,feb_NDSI,mar_NDSI,apr_NDSI,may_NDSI,jun_NDSI,jul_NDSI,aug_NDSI,sept_NDSI,oct_NDSI,nov_NDSI,dec_NDSI) %>%
  rename(NDSI = c.0.209958436213992..0.198818340611354...0.0420229007633588..)
snowiest_month <- max(snow_NDSI)
```
The snowiest month is January, with an average NDSI of `r snowiest_month`. 

## Bonus Question: Redo all problems with `spread` and `gather` using modern tidyverse syntax. 

## Bonus Question: Use Climage Engine to pull the same data for the assignment, but updated with 2020/2021 data.




